---
title: "Statistical Inference Course Project"
author: "Omer Kara"
date: "Friday, January 15, 2016"
output:
  html_document:
    keep_md: yes
    theme: readable
  pdf_document:
    fig_height: 5
subtitle: 'Part 1: Simulation Exercises'
---
Source code for this entire report can be found [here](https://github.com/omerkara/Coursera_Statistical-Inference).

```{r echo = FALSE, warning = FALSE}
Load.Install <- function(package_names) {
    is_installed <- function(mypkg) is.element(mypkg, utils::installed.packages()[ ,1])
    for (package_name in package_names) {
        if (!is_installed(package_name)) {
            utils::install.packages(package_name, dependencies = TRUE)
        }
        suppressMessages(library(package_name, character.only = TRUE, quietly = TRUE, verbose = FALSE))
    }
}
Load.Install(c("ggplot2", "nortest"))
```

## Overview
The exponential distribution (a.k.a. negative exponential distribution) is the probability distribution that describes the time between events in a Poisson process, i.e. a process in which events occur continuously and independently at a constant average rate.[^1] It is a special case of the gamma distribution. The exponential distribution can be simulated in R with `rexp(n, lambda)` where `lambda` such that $\lambda > 0$ is the parameter of the distribution, often called the rate parameter. The distribution is supported on the interval $[{0, \infty})$. If a random variable X has this distribution, we write $X \sim Exp(\lambda)$. Note that the mean and standard deviation of exponential distribution is $1/\lambda$. For more information about the moments of exponential distribution please see [Exponential Distribution Wiki Page](https://en.wikipedia.org/wiki/Exponential_distribution).

For this simulation exercie, $\lambda=0.2$ is given. In this simulation, we investigate the distribution of averages of 40 observations sampled from exponential distribution with $\lambda=0.2$ with 1000 simulations.  

## Simulation
```{r echo = TRUE}
set.seed(1)
lambda <- 0.2
obs <- 40
sim <- 1000
data <- data.frame(means = apply(matrix(rexp(sim * obs, rate = lambda), sim), 1, mean))
```

## Theoretical vs. Empirical Distribution
The central limit theorem (CLT) states that averages of sufficiently larger number of independently and identically distributed observations, each with a well-defined expected value and well-defined variance, will be asymptotically normally distributed, regardless of the underlying distribution. Therefore, following the theory, we are expecting to see our simulation data is distributed as $\overline{X} \sim \mathcal{N}(\mu,\sigma^2/n)$. More specifically, in our case the theoretical distribution is as follows $\overline{X} \sim \mathcal{N}(\lambda^{-1},\lambda^{-2}/n)$. 

The following results clearly shows that our simulation coincides with the theosry very well since the moments of empirical distribution is very close to the theoretical counterparts.

```{r echo = TRUE}
## Theoretical vs. Empirical mean.
mu <- c(1/lambda, mean(data$means)) 
mu
## Theoretical vs. Empirical variance.
var <- c(1/((lambda ^ 2)*obs), var(data$means))
var
## Theoretical standard deviation vs. Empirical standard error.
sd <- c(sqrt(1/((lambda ^ 2)*obs)), sd(data$means))
sd
```

## Empirical and Theoretical Distributions
The figure below shows the theoretical distribution with `red` lines and empirical distribution with `black` lines.[^2] As it is clearly shown, the empirical distribution is asymptotically normally distributed. Both densities are almost overlapped, also means and spread (variance) are very close to each other.
```{r echo = FALSE}
g <- ggplot(data = data, aes(x = means)) ## Main plot.
g <- g + labs(title = expression("Distribution of Sample Means Drawn from Exponential Distribution with" ~ lambda ~ "= 0.2")) + labs(x = expression("Sample Means (" * hat(mu) * ")"), y = "Density") + scale_x_continuous(breaks = round(seq(min(data$means), max(data$means), by = 1))) ## Labels and title.
g <- g + geom_histogram(aes(y = ..density..), fill = "black", alpha = 0.5, binwidth = 0.15, colour = "black") + geom_density(size = 0.5) ## Histogram and density of the empirical distribution. 
g <- g + geom_vline(xintercept = mean(data$means), color = "black", size = 1) ## Mean of the empirical distribution.
g <- g + stat_function(fun = dnorm, args = list(mean = 1/lambda, sd = (1/lambda)/sqrt(obs)), geom = "line", size = 0.5, col = "red") ## Density of theoretical distribution. 
g <- g + geom_vline(xintercept = 1/lambda, color = "red", size = 0.5) ## Mean of the theoretical distribution.
g
```


## Formal Normality Tests
The Shapiro-Wilk and Anderson-Darling normality test suggests that the empirical distribution is in fact normal. Both tests fail to reject the null hypothesis which indicates the data is normal.
```{r echo = TRUE}
shapiro.test(data$means) ## Normality check with Shapiro-Wilk test.
ad.test(data$means) ## Normality check with Anderson-Darling test.
```

[^1]: https://en.wikipedia.org/wiki/Exponential_distribution
[^2]: Veritcal lines are the respective means.

